---
title: "Data Science Salary Analysis"
author: "Malcolm Brea"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(tidyverse)
```

---

## A.) Restatement of the Question

**Original Question (paraphrased):**
Our CEO wants to hire a senior data scientist (potentially the first of a future team) and needs to know what salary range to offer in order to attract and retain top talent in a highly competitive, rapidly escalating market. The role may be offshore or U.S.-based, and the pay must reflect both current market conditions and the candidate's future leadership potential.

**Alternative framings of the question:**

1. What is the market rate for an experienced, team-leading data scientist in the United States vs. offshore?
2. If we want to be in the top quartile of employers competing for senior DS talent, what should our compensation floor be?
3. How does geography (U.S. vs. international) affect data scientist salaries at the mid-to-senior level, and what premium must we pay for domestic talent?
4. What is the minimum salary needed to recruit someone capable of growing into a data science manager within two to three years?

**What could be added to or removed from the question:**

- Add: Benefits, equity, remote work flexibility, and total compensation context. Salary alone does not tell the full story.
- Add: Industry-specific norms (tech vs. finance vs. healthcare) could shift the benchmark significantly.
- Remove: If budget is fixed, reframe from "what should we pay?" to "what seniority level can we afford?"
- Add: A timeline element. Salaries are rising, the range today may be obsolete in 12 months.

---

## B.) Data Loading and Preparation

### B.1 — Load the Data

```{r load_data}
df <- read.csv("project_1_data.csv", stringsAsFactors = FALSE)

# Drop the unnamed index column
if ("X" %in% colnames(df)) df <- df[, colnames(df) != "X"]

cat("Dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
head(df, 3)
```

### B.2 — Set Variable Types as Factors

```{r factor_conversion}
df$work_year <- as.factor(df$work_year)

df$experience_level <- factor(df$experience_level,
                              levels = c("EN", "MI", "SE", "EX"),
                              labels = c("Entry", "Mid", "Senior", "Executive"))

df$employment_type <- factor(df$employment_type,
                             levels = c("FT", "PT", "CT", "FL"),
                             labels = c("Full-Time", "Part-Time", "Contract", "Freelance"))

df$job_title          <- as.factor(df$job_title)
df$salary_currency    <- as.factor(df$salary_currency)
df$employee_residence <- as.factor(df$employee_residence)

df$remote_ratio <- factor(df$remote_ratio,
                          levels = c(0, 50, 100),
                          labels = c("On-site", "Hybrid", "Remote"))

df$company_location <- as.factor(df$company_location)

df$company_size <- factor(df$company_size,
                          levels = c("S", "M", "L"),
                          labels = c("Small", "Medium", "Large"))

str(df)
```

### B.3 — Basic Exploratory Data Analysis

#### Summary Statistics

```{r eda_summary}
summary(df)
```

**Commentary on each variable:**

- **work_year**: The dataset spans 2020–2022. Records roughly triple from 2020 to 2022, reflecting rapid growth in the data science labor market and the proliferation of job postings post-pandemic.
- **experience_level**: Senior roles dominate (n=280, ~46%), consistent with market demand for proven practitioners over entry-level hires. Executive roles are rare (n=26), typical for an emerging field.
- **employment_type**: Full-time positions make up 97% of records — the vast majority of DS hiring is for permanent roles, not contract or freelance.
- **job_title**: 50+ unique titles in the dataset. "Data Scientist" and "Data Engineer" are most common, but salary varies dramatically by title and seniority.
- **salary / salary_in_usd**: The raw salary column reflects local currency. `salary_in_usd` normalizes everything to USD — this is the appropriate column for all cross-country comparisons.
- **salary_currency**: 17 currencies present; USD dominates. Converting to `salary_in_usd` is essential.
- **employee_residence**: Closely mirrors `company_location` — most employees work in the same country as the company.
- **remote_ratio**: Remote-only (100%) is the plurality mode (~64%), a post-pandemic norm in data science.
- **company_location**: U.S. companies account for ~58% of the dataset (n=355), making it the dominant market.
- **company_size**: Large companies (L) are the biggest employer segment and typically pay the most.

#### Additional Tabular Summary — Salary by Experience Level

```{r eda_table}
salary_by_exp <- df %>%
  group_by(experience_level) %>%
  summarise(
    Count      = n(),
    Min_USD    = dollar(min(salary_in_usd)),
    Q1_USD     = dollar(quantile(salary_in_usd, 0.25)),
    Median_USD = dollar(median(salary_in_usd)),
    Mean_USD   = dollar(mean(salary_in_usd)),
    Q3_USD     = dollar(quantile(salary_in_usd, 0.75)),
    Max_USD    = dollar(max(salary_in_usd))
  )
kable(salary_by_exp, caption = "Salary Distribution by Experience Level (USD) — All Roles")
```

---

#### EDA Plots — One Per Variable

```{r plot_work_year, fig.width=7, fig.height=4}
ggplot(df, aes(x = work_year, fill = work_year)) +
  geom_bar(show.legend = FALSE) +
  labs(title = "Number of Records by Work Year",
       subtitle = "Dataset grows year-over-year as the field expands",
       x = "Year", y = "Count") +
  scale_fill_brewer(palette = "Blues") + theme_minimal()
```

```{r plot_experience, fig.width=7, fig.height=4}
ggplot(df, aes(x = experience_level, y = salary_in_usd, fill = experience_level)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.4) +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(title = "Salary by Experience Level",
       subtitle = "Clear step-up at each level; Executive variance is large due to small n",
       x = "Experience", y = "Salary (USD)") +
  scale_fill_brewer(palette = "Greens") + theme_minimal()
```

```{r plot_employment, fig.width=7, fig.height=4}
ggplot(df, aes(x = employment_type, fill = employment_type)) +
  geom_bar(show.legend = FALSE) +
  labs(title = "Employment Type Distribution",
       subtitle = "Full-time roles overwhelmingly dominate (97%)",
       x = NULL, y = "Count") +
  scale_fill_brewer(palette = "Oranges") + theme_minimal()
```

```{r plot_job_title, fig.width=9, fig.height=5}
df %>%
  group_by(job_title) %>%
  summarise(med_sal = median(salary_in_usd), n = n()) %>%
  filter(n >= 5) %>%
  slice_max(order_by = med_sal, n = 12) %>%
  ggplot(aes(x = reorder(job_title, med_sal), y = med_sal, fill = med_sal)) +
  geom_col(show.legend = FALSE) + coord_flip() +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  scale_fill_gradient(low = "#A8D8EA", high = "#065A82") +
  labs(title = "Median Salary by Job Title (titles with n ≥ 5)",
       x = NULL, y = "Median Salary (USD)") + theme_minimal()
```

```{r plot_salary_dist, fig.width=7, fig.height=4}
ggplot(df, aes(x = salary_in_usd)) +
  geom_histogram(fill = "#065A82", bins = 40, alpha = 0.8) +
  scale_x_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(title = "Overall Salary Distribution (USD)",
       subtitle = "Right-skewed: bulk between $50K–$200K, with high-end outliers",
       x = "Salary (USD)", y = "Count") +
  theme_minimal()
```

```{r plot_remote, fig.width=7, fig.height=4}
ggplot(df, aes(x = remote_ratio, y = salary_in_usd, fill = remote_ratio)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.4) +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(title = "Salary by Remote Work Status",
       subtitle = "Remote roles skew slightly higher — driven by U.S. company dominance",
       x = NULL, y = "Salary (USD)") +
  scale_fill_brewer(palette = "Purples") + theme_minimal()
```

```{r plot_location, fig.width=7, fig.height=5}
df %>%
  group_by(company_location) %>%
  summarise(med = median(salary_in_usd), n = n()) %>%
  filter(n >= 10) %>%
  ggplot(aes(x = reorder(company_location, med), y = med, fill = med)) +
  geom_col(show.legend = FALSE) + coord_flip() +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  scale_fill_gradient(low = "#B2DFDB", high = "#00695C") +
  labs(title = "Median Salary by Company Location (n ≥ 10)",
       x = "Country Code", y = "Median Salary (USD)") + theme_minimal()
```

```{r plot_company_size, fig.width=7, fig.height=4}
ggplot(df, aes(x = company_size, y = salary_in_usd, fill = company_size)) +
  geom_violin(show.legend = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  labs(title = "Salary Distribution by Company Size",
       subtitle = "Large companies pay more and have less variance",
       x = NULL, y = "Salary (USD)") +
  scale_fill_brewer(palette = "Set2") + theme_minimal()
```

---

## C.) Tables and Graphs Answering the CEO's Question

### C.1 — Table 1: US vs. Offshore Salary by Experience Level (Full-Time)

```{r ceo_table1}
us_vs_offshore <- df %>%
  filter(employment_type == "Full-Time") %>%
  mutate(region = ifelse(company_location == "US", "United States", "Offshore")) %>%
  group_by(region, experience_level) %>%
  summarise(
    n          = n(),
    Median_USD = dollar(median(salary_in_usd)),
    Mean_USD   = dollar(mean(salary_in_usd)),
    P25_USD    = dollar(quantile(salary_in_usd, 0.25)),
    P75_USD    = dollar(quantile(salary_in_usd, 0.75)),
    .groups    = "drop"
  )
kable(us_vs_offshore,
      caption = "Full-Time Salary Benchmarks: U.S. vs. Offshore by Experience Level (USD)")
```

### C.2 — Table 2: Senior & Executive US Salary Benchmarks by Title

```{r ceo_table2}
us_senior <- df %>%
  filter(company_location == "US",
         employment_type  == "Full-Time",
         experience_level %in% c("Senior", "Executive")) %>%
  group_by(job_title) %>%
  summarise(
    n          = n(),
    Min_USD    = dollar(min(salary_in_usd)),
    Median_USD = dollar(median(salary_in_usd)),
    P75_USD    = dollar(quantile(salary_in_usd, 0.75)),
    Max_USD    = dollar(max(salary_in_usd)),
    .groups    = "drop"
  ) %>%
  filter(n >= 3) %>%
  arrange(desc(parse_number(Median_USD)))
kable(us_senior,
      caption = "Senior/Executive Data Science Salaries — U.S. Full-Time Roles (n ≥ 3)")
```

### C.3 — Graph 1: Salary Trend Over Time (US vs. Offshore, Full-Time)

```{r ceo_graph1, fig.width=9, fig.height=5}
trend <- df %>%
  filter(employment_type == "Full-Time") %>%
  mutate(region = ifelse(company_location == "US", "United States", "Offshore")) %>%
  group_by(work_year, region) %>%
  summarise(med_sal = median(salary_in_usd), .groups = "drop")

ggplot(trend, aes(x = work_year, y = med_sal, color = region, group = region)) +
  geom_line(linewidth = 1.8) +
  geom_point(size = 4) +
  geom_text(aes(label = dollar(med_sal, scale = 1e-3, suffix = "K")),
            vjust = -.5, fontface = "bold", size = 3.5) +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K"), limits = c(0, NA)) +
  scale_color_manual(values = c("United States" = "#065A82", "Offshore" = "#02C39A")) +
  labs(title = "Data Scientist Salaries Are Rising — Fast",
       subtitle = "Median full-time salary by year | U.S. vs. Offshore",
       x = "Year", y = "Median Annual Salary (USD)", color = "") +
  theme_minimal(base_size = 13) + theme(legend.position = "bottom")
```

### C.4 — Graph 2: Salary Range by Experience — US vs. Offshore

```{r ceo_graph2, fig.width=9, fig.height=5}
df %>%
  filter(employment_type == "Full-Time") %>%
  mutate(region = ifelse(company_location == "US", "United States", "Offshore")) %>%
  ggplot(aes(x = experience_level, y = salary_in_usd, fill = region)) +
  geom_boxplot(outlier.alpha = 0.3, position = position_dodge(0.8)) +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  scale_fill_manual(values = c("United States" = "#065A82", "Offshore" = "#02C39A")) +
  labs(title = "The U.S. Premium Is Real — and Grows with Seniority",
       subtitle = "Full-time positions | Boxes = 25th–75th percentile",
       x = "Experience Level", y = "Annual Salary (USD)", fill = "") +
  theme_minimal(base_size = 13) + theme(legend.position = "bottom")
```

---

## Story Summary

The data are unambiguous. Based on 607 real data science job records from 2020–2022, U.S.-based senior data scientists are expensive and getting more expensive every year, with median salaries rising from \$114,000 in 2020 to \$140,000 in 2022, a 23% jump in just two years. Offshore senior talent runs at roughly half that cost, with a median of \$88,000 and an interquartile range of \$62K–\$114K. The gap between the two markets is not shrinking — it is widening. For a small but fast-growing company seeking its first senior data scientist, someone who can both deliver results today and build a team tomorrow, our recommendation is to budget \$140,000-\$180,000 for a U.S.-based hire (top quartile positioning) or \$75,000-\$100,000 for an offshore equivalent. Hiring now is the right move: every quarter of delay means a higher asking price in a market where supply has not caught up to demand.
